% Abstract (150-250 words, self-contained, one paragraph, no equations/tables/refs)
Failure Mode and Effects Analysis (FMEA) is a cornerstone of reliability engineering for nanosatellite missions. Despite its established utility, the method presents a well-known structural limitation: the subjectivity inherent in the manual assignment of Severity (S), Occurrence (O), and Detection (D) indices, whose product yields the Risk Priority Number (RPN). This reliance on expert judgment compromises reproducibility and introduces inter-rater variability, particularly in university-led programs with teams in training. To address this problem, this work proposes an Explainable Artificial Intelligence framework based on the XGBoost algorithm to automate and calibrate RPN estimation in CubeSat missions. The model was trained on a structured dataset of 80~failure modes compiled from CubeSat subsystem analyses and complemented by data from the specialized literature. The approach was directly compared against conventional expert-based assessment, and XGBoost achieved a coefficient of determination of 0.94 on the test set, outperforming both Linear Regression and Random Forest baselines. SHapley Additive Explanations analysis was applied to ensure interpretability, revealing that Detection is the dominant risk driver in communication and structural subsystems, while Severity dominates in attitude control systems. Additionally, hierarchical clustering classified the failure modes into three actionable risk groups, providing structured support for corrective action prioritization. The results indicate that the proposed framework can reduce inter-rater variability, increase process traceability, and make FMEA analysis reproducible in space programs with limited resources.