\section{Related Work}
\label{sec:related_work}

This section reviews the state of the art across four interconnected domains: CubeSat mission reliability, machine learning applied to FMEA, anomaly detection in satellite telemetry, and explainable AI in safety-critical systems. The identified gap in the literature is discussed at the end.

\subsection{Reliability and Failure Analysis in CubeSat Missions}
\label{subsec:cubesat_reliability}

The systematic study of CubeSat failures gained significant scientific momentum with the statistical survey conducted by Villela et al.~\cite{villela2019towards}, who compiled and analyzed data from hundreds of missions up to 2019, establishing a quantitative baseline for success rates by subsystem. Their work revealed that communication and power subsystems concentrate the majority of documented functional failures---a finding that has since become a reference for calibrating occurrence scales in subsequent FMEA analyses.

Bouwmeester, Menicucci, and Gill~\cite{bouwmeester2022improving} investigated the trade-off between subsystem redundancy and improved testing as strategies to enhance CubeSat reliability. Using Kaplan--Meier estimators and Bayesian inference on historical failure data, they concluded that investing in improved testing tends to be more effective than adding redundancy, particularly for short-duration missions. This conclusion has direct implications for the definition of corrective actions in FMEA analyses, reinforcing the importance of accurately identifying the subsystem and mission phase associated with each failure mode.

At the component level, Albalooshi, Jallad, and Marpu~\cite{albalooshi2023fault} performed a detailed fault analysis of the I2C communication bus in nanosatellite missions, identifying critical risks to mission success and proposing mitigation techniques. Zeedan and Khattab~\cite{zeedan2023cubesat} provided a comprehensive review of CubeSat communication subsystem architectures, highlighting the challenges and failure-prone aspects of on-board transceivers. In the domain of CubeSat testing, Gutierrez et al.~\cite{gutierrez2021fuzz} demonstrated the application of systematic fuzz testing techniques on nanosatellite flight software, while Sureda et al.~\cite{sureda2021helix} reported on the design and testing of deployment mechanisms for CubeSat antennas, a subsystem where deployment failures represent a significant risk factor. These works collectively highlight that single-point-of-failure components in CubeSat architectures can lead to catastrophic mission outcomes, further motivating the need for systematic risk prioritization methods such as the one proposed in this work.

\subsection{Machine Learning for RPN Estimation in FMEA}
\label{subsec:ml_rpn}

The automation of the FMEA process through machine learning has been investigated in industrial contexts with notable results. Al-Refaie et al.~\cite{alrefaie2020enhancing} proposed the application of AutoML for automatic classification of failure modes and RPN estimation without human intervention, demonstrating the conceptual feasibility of the approach. More recently, an integrated framework based on machine learning was applied to failure prioritization and process risk analysis in manufacturing, achieving high predictive performance in RPN estimation~\cite{ref4_ml_fmea2026}. These works established that gradient boosting models are particularly well-suited for capturing the nonlinear relationships between risk variables in tabular FMEA datasets.

The utilization of Large Language Models (LLMs) for FMEA automation has also begun to emerge in the literature. Thomas~\cite{thomas2023revolutionizing} explored the use of ChatGPT for automatic generation of FMEA tables from textual component descriptions, identifying both the potential and the limitations of this approach in terms of consistency and traceability. While LLM-based approaches offer flexibility in text generation, they lack the quantitative rigor and reproducibility of supervised regression models, which remain preferable for numerical RPN estimation.

Several works published in IEEE Access and IEEE Transactions on Reliability have addressed the limitations of the conventional RPN calculation. Liu et al.~\cite{amwrpn2018} proposed the AMWRPN model, which weights risk factors using ambiguity measures based on Dempster--Shafer evidence theory, addressing the uncertainty inherent in expert assessments. The same research group earlier proposed an integrated multiple criteria decision making approach for FMEA under uncertainty~\cite{liu2016fmea_uncertainty}. Huang et al.~\cite{geomean_fmea2019} introduced a geometric mean FMEA method based on information quality, using fuzzy distributions to capture expert opinion more faithfully. Jiang, Zhang, and Deng~\cite{jiang2019fuzzy_fmea} proposed a fuzzy evidential reasoning-based FMEA method that combines fuzzy sets with Dempster--Shafer theory to handle the vagueness in expert assessments. Zhang and Chu~\cite{aadfmea2023} proposed an approach based on avoidance of aggregation discrepancy for group-based FMEA evaluation. While these methods improve upon conventional RPN through mathematical and statistical reformulations, none of them leverages machine learning or provides the per-subsystem interpretability that explainable AI techniques can offer.

\subsection{Anomaly Detection in Satellite Telemetry with Machine Learning}
\label{subsec:anomaly_detection}

Machine learning techniques for anomaly detection have been extensively surveyed by Nassif et al.~\cite{nassif2021anomaly}, who provided a systematic review of supervised, unsupervised, and semi-supervised approaches across multiple application domains. In the specific context of satellite telemetry analysis, Cespedes et al.~\cite{cespedes2022performance} conducted a comparative evaluation of five machine learning algorithms for anomaly detection in CubeSat solar panels from the BIRDS-3 and BIRDS-4 constellations, considering not only predictive accuracy but also memory and energy consumption in resource-constrained embedded environments. The study concluded that linear models offer the best trade-off between performance and on-board inference feasibility---a result relevant for future extensions of this work toward in-orbit monitoring.

A significant advancement in this direction is represented by the OPS-SAT Anomaly Detection benchmark introduced by Ruszczak et al.~\cite{ruszczak2025opssat}, published in Scientific Data. This constitutes the first publicly available dataset of real CubeSat telemetry---from ESA's OPS-SAT, a 3U satellite operated for over four years---annotated and accompanied by baseline results from 30 supervised and unsupervised algorithms. This resource represents a concrete opportunity for external validation and transfer learning in future works that seek to extend the proposed framework to the operational phase.

\subsection{Explainable AI in Safety-Critical Engineering Systems}
\label{subsec:xai}

The adoption of Explainable AI (XAI) techniques in safety-critical systems has grown substantially, driven by both regulatory compliance requirements and the demand for transparency in decision-making~\cite{adadi2018xai,ortigossa2024xai}. Chen et al.~\cite{chen2024interpretable} provided a comprehensive survey of interpretable fault diagnosis methods, emphasizing the role of interpretability in enhancing the reliability of ML-based systems. Among the available techniques, SHAP (SHapley Additive Explanations), derived from cooperative game theory, stands out as the most widely adopted method due to its ability to generate both local and global explanations in a mathematically consistent manner. SHAP is particularly effective when combined with tree-based ensemble models such as XGBoost, for which efficient exact computation algorithms exist~\cite{xai_xgboost_shap2025}.

Recent applications of XGBoost with SHAP interpretability have demonstrated the value of this combination across diverse engineering domains. In energy forecasting, Wang et al.~\cite{wang2025xgboostshap} proposed an XGBoost-SHAP model for energy demand prediction with feature selection, achieving strong predictive accuracy and transparent feature importance quantification. In predictive maintenance, Gawde et al.~\cite{gawde2024explainable} applied LIME, SHAP, PDP, and ICE to rotating machinery fault diagnosis, demonstrating that XAI methods substantially enhance practitioner trust in model outputs. Cummins et al.~\cite{cummins2024explainable} provided a comprehensive survey of explainable predictive maintenance methods, identifying SHAP as the dominant interpretability technique in current practice. Alam et al.~\cite{alam2024sxad} proposed SXAD, a SHAP-based anomaly detection framework that leverages Shapley values for transparent identification of anomalous patterns. In the manufacturing domain, Akyaz and Eng{\i}n~\cite{akyaz2024predictive} demonstrated the effectiveness of machine learning-based predictive maintenance systems in industrial settings. However, applications in the aerospace reliability domain, particularly for FMEA automation in CubeSat missions, remain absent from the literature.

\subsection{Identified Gap in the Literature}
\label{subsec:gap}

The review of the state of the art reveals a clear gap that this work aims to fill. While the automation of FMEA using machine learning has been demonstrated in industrial manufacturing contexts, and while anomaly detection in satellite telemetry using ML is expanding rapidly, no existing work combines XGBoost-based RPN prediction with SHAP explainability analysis specifically applied to CubeSat failure modes. Furthermore, none of the reviewed FMEA improvement methods published in IEEE Access~\cite{amwrpn2018,geomean_fmea2019,aadfmea2023} employs machine learning, and none provides per-subsystem interpretability through explainable AI techniques.

This article fills this gap by proposing, implementing, and validating an end-to-end XAI framework for CubeSat FMEA that integrates data-driven RPN estimation, SHAP-based per-subsystem risk driver identification, and hierarchical clustering for actionable risk group classification.
